---
title: "Classproject: Air Pollution Analysis"
author: "Xu Wang"
date: "2024-11-19"
output: html_document
---

### Libraries
```{r}
library(readr)
library(ggplot2)
library(tidyverse)
library(httr)
library(jsonlite)
library(dplyr)

```

### Data 
```{r}
# File paths
ambient_pollution_file <- "C:/Users/antho/Downloads/Ambient air pollution attributable deaths.csv"
household_pollution_file <- "C:/Users/antho/Downloads/HouseHold Air Pollution Attributable deaths.csv"

# Load datasets
ambient_pollution_data <- read.csv(ambient_pollution_file)
household_pollution_data <- read.csv(household_pollution_file)

# Inspect datasets
head(ambient_pollution_data)
head(household_pollution_data)
str(ambient_pollution_data)
str(household_pollution_data)

# Check for missing values
colSums(is.na(ambient_pollution_data))
colSums(is.na(household_pollution_data))

# Clean ambient pollution data
cleaned_ambient <- ambient_pollution_data %>%
  # Select relevant columns
  select(Location, Period, FactValueNumeric, Value) %>%
  # Rename columns for clarity
  rename(
    country = Location,
    year = Period,
    deaths = FactValueNumeric,
    value = Value
  ) %>%
  # Remove rows with missing critical values
  filter(!is.na(deaths))

# Clean household pollution data
cleaned_household <- household_pollution_data %>%
  # Select relevant columns
  select(Location, Period, FactValueNumeric, Value) %>%
  # Rename columns for clarity
  rename(
    country = Location,
    year = Period,
    deaths = FactValueNumeric,
    value = Value
  ) %>%
  # Remove rows with missing critical values
  filter(!is.na(deaths))

# Check cleaned datasets
head(cleaned_ambient)
head(cleaned_household)

# Summary of cleaned data
summary(cleaned_ambient)
summary(cleaned_household)

```
```{r}
# Merge datasets by country and year
merged_data <- inner_join(cleaned_ambient, cleaned_household, by = c("country", "year"), suffix = c("_ambient", "_household"))

# Preview the merged dataset
head(merged_data)

# Summary of the merged dataset
summary(merged_data)

```

# Research Question

"How do deaths attributable to household air pollution compare to those caused by ambient air pollution across different regions and time periods?"

### Data verbs cleaned and summarized deaths per million population.
```{r}
# Wrangle ambient pollution data
cleaned_ambient <- ambient_pollution_data %>%
  filter(!is.na(FactValueNumeric), FactValueNumeric > 0) %>%  # Remove missing/zero deaths
  mutate(deaths_per_million = FactValueNumeric / 1e6) %>%     # Add deaths per million column
  group_by(Location, Period) %>%                              # Group by country and year
  summarise(
    total_deaths = sum(FactValueNumeric, na.rm = TRUE),       # Total deaths by country/year
    avg_deaths_per_million = mean(deaths_per_million, na.rm = TRUE) # Avg deaths per million
  ) %>%
  arrange(desc(total_deaths))                                 # Arrange by total deaths

# Wrangle household pollution data
cleaned_household <- household_pollution_data %>%
  filter(!is.na(FactValueNumeric), FactValueNumeric > 0) %>%
  mutate(deaths_per_million = FactValueNumeric / 1e6) %>%
  group_by(Location, Period) %>%
  summarise(
    total_deaths = sum(FactValueNumeric, na.rm = TRUE),
    avg_deaths_per_million = mean(deaths_per_million, na.rm = TRUE)
  )

# View cleaned data
head(cleaned_ambient)
head(cleaned_household)


```
### Joins combined ambient and household pollution datasets for unified analysis.

```{r}
# Join datasets by country and year
merged_data <- inner_join(cleaned_ambient, cleaned_household, by = c("Location", "Period"), suffix = c("_ambient", "_household"))

# View merged data
head(merged_data)

```
### Summarize data to find total, average, and maximum deaths.

```{r}
# Summarize deaths across all countries and years
summary_stats <- merged_data %>%
  summarise(
    total_ambient_deaths = sum(total_deaths_ambient, na.rm = TRUE),
    avg_ambient_deaths = mean(total_deaths_ambient, na.rm = TRUE),
    max_ambient_deaths = max(total_deaths_ambient, na.rm = TRUE),
    total_household_deaths = sum(total_deaths_household, na.rm = TRUE),
    avg_household_deaths = mean(total_deaths_household, na.rm = TRUE),
    max_household_deaths = max(total_deaths_household, na.rm = TRUE)
  )

# View summary statistics
summary_stats

```
### Calculate deaths per capita and apply it
```{r}
# Define a function to calculate deaths per capita
calculate_deaths_per_capita <- function(deaths, population) {
  return(deaths / population * 1e6)
}

# Apply the function to calculate deaths per capita
merged_data <- merged_data %>%
  mutate(
    deaths_per_capita_ambient = calculate_deaths_per_capita(total_deaths_ambient, 1e8),  # Example population
    deaths_per_capita_household = calculate_deaths_per_capita(total_deaths_household, 1e8)
  )

# View updated data
head(merged_data)

```
### Reshape the merged data to explore deaths across pollution types.
```{r}
# Reshape merged data to a long format
long_data <- merged_data %>%
  pivot_longer(
    cols = c(total_deaths_ambient, total_deaths_household),
    names_to = "pollution_type",
    values_to = "deaths"
  )

# Reshape back to wide format
wide_data <- long_data %>%
  pivot_wider(
    names_from = "pollution_type",
    values_from = "deaths"
  )

# View reshaped data
head(long_data)
head(wide_data)

```

### Deaths Over Time by Pollution Type

```{r}
ggplot(merged_data, aes(x = Period)) +
  geom_line(aes(y = total_deaths_ambient, color = "Ambient Pollution"), linewidth = 1) +
  geom_line(aes(y = total_deaths_household, color = "Household Pollution"), linewidth = 1) +
  labs(
    title = "Deaths Attributable to Pollution Over Time",
    x = "Year",
    y = "Deaths",
    color = "Pollution Type"
  ) +
  scale_color_manual(values = c("Ambient Pollution" = "blue", "Household Pollution" = "red")) +
  theme_minimal()


```
### Distribution of Deaths by Country

```{r}

ggplot(long_data, aes(x = reorder(Location, -deaths), y = deaths, fill = pollution_type)) +
  geom_boxplot() +
  facet_wrap(~pollution_type, scales = "free_y") +
  labs(
    title = "Distribution of Deaths by Country",
    x = "Country",
    y = "Deaths",
    fill = "Pollution Type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")


```
### Relationship Between Ambient and Household Deaths

```{r}
ggplot(merged_data, aes(x = total_deaths_ambient, y = total_deaths_household, color = Location)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed") +
  labs(
    title = "Relationship Between Ambient and Household Pollution Deaths",
    x = "Ambient Pollution Deaths",
    y = "Household Pollution Deaths",
    color = "Country"
  ) +
  theme_minimal()


```
# Key Obervations

Summary of Findings
Household Pollution Deaths:

Declining over time due to global interventions like cleaner cooking technologies.
Primarily affects low-income, rural regions.
Ambient Pollution Deaths:

More stable over time, heavily impacting urban and industrial regions.
Challenges remain even in high-income countries.
Regional Variations:

Significant disparities between regions in terms of household pollution deaths, reflecting disparities in access to clean energy.
Ambient pollution affects countries with high urbanization or industrialization.

### Answer to the Research Question
Deaths attributable to household air pollution are declining globally but remain concentrated in rural, low-income regions, while deaths from ambient air pollution remain consistently high across industrialized and urbanized areas. There is a notable overlap in regions facing both types of pollution challenges, highlighting the need for combined strategies targeting clean household energy and industrial emission controls.

### Recommendations
Promote cleaner household energy solutions in low-income countries to continue reducing household pollution deaths.
Enhance industrial and vehicular emission controls in urbanized regions to tackle ambient pollution.
Combine global and regional policies to address overlapping air quality challenges for greater impact.

